ProbabilityBoundsAnalysis.jl:\\ an arithmetic of probabilities and sets of probabilities

### 2021-09-09 paper/paper.tex
adds p-box theory and fig1

% JuliaCon proceedings template
\documentclass{juliacon}
\setcounter{page}{1}
\usepackage{caption}
\usepackage{csquotes}
%\bibliographystyle{juliacon}

@@

An arithmetic of probability distributions has held a long interest among mathematicians and engineers:
\begin{displayquote}
  A question asked by Kolmogorov, \\
  answered for the sum by Marakov, \\
  partially answered by Sklar and Frank (for which the copula was invented),\\
  made algorithmically available by Williamson, \\
  and generalised by others.
\end{displayquote}

@@

Imprecise probabilities is effectively a generalisation of probability theory where uncertainty can be expressed about the probability measure. This 
is particularly relevant when information is scarce, unreliable, vague, conflicting or imprecise. In such cases defining a unique probability distribution is difficult. P-boxes are one of many ways to describe a set of distributions, others include: Dempster-Shafer structures \cite{dempster2008upper,shafer1976mathematical}, random sets \cite{molchanov2005theory}, possibility distributions \cite{zadeh1978fuzzy,dubois1988possibility, hose2019possibilistic} and lower previsions \cite{troffaes2014lower}. These structures were discovered independently, but are often synonymous and can be translated from one to another, with different degrees of generality. Imprecise probabilities links all these theories into one. For a comprehensive overview of the theory, and a for a formal description of uncertainty and information in terms of these structures, \cite{klir2013uncertainty} is recommended. In that sense PBA is a part of imprecise probabilities but provides a framework for computing with p-boxes. In this section we review the main elements of the PBA.

@@

A probability box defines a set of distributions with the following three constrains:

@@

\subsubsection{Partial moment information} %\hfill \break
%It is well known that a random variables distribution function cannot be

@@

\section{Operations on p-boxes}

@@

The long term goal of such a framework is to create a fully uncertain programming language, where any computer variable may be represented as an interval, distribution, p-box or other uncertain quantity. Such a framework would allow for uncertain extensions of deterministic functions to be computed in an automatic, rigorous and tight fashion. In this section we argue why Julia is an ideal target language for such a framework, and discuss the remaining theoretical tasks required to make such a goal a reality.

>>>>

% JuliaCon proceedings template
\documentclass{juliacon}
\setcounter{page}{1}
\usepackage{caption}
\usepackage{csquotes}
\usepackage{amsmath}
%\bibliographystyle{juliacon}

@@

An arithmetic of probability distributions has held a long interest among mathematicians and scientists:
\begin{displayquote}
  A question asked by Kolmogorov, \\
  answered for the sum by Marakov, \\
  partially answered by Sklar and Frank (for which the copula was invented),\\
  made algorithmically available by Williamson, \\
  and generalised by others.
\end{displayquote}

@@

Imprecise probabilities is effectively a generalisation of probability theory where uncertainty can be expressed about the probability measure. This 
is particularly relevant when information is scarce, unreliable, vague, conflicting or imprecise. In such cases defining a unique probability distribution is difficult. P-boxes are one of many ways to describe a set of distributions, others include: Dempster-Shafer structures \cite{dempster2008upper,shafer1976mathematical}, random sets \cite{molchanov2005theory}, possibility distributions \cite{zadeh1978fuzzy,dubois1988possibility, hose2019possibilistic} and lower previsions \cite{troffaes2014lower}. These structures were discovered independently, but are often synonymous and can be translated from one to another, with different degrees of generality. Imprecise probabilities links all these theories into one. For a comprehensive overview of the theory, and a for a formal description of uncertainty and information in terms of these structures, \cite{klir2013uncertainty} is recommended. In that sense PBA is a part of imprecise probabilities but provides a framework for computing with p-boxes.

@@

A probability box defines a set of distributions with the following three constrains: (1) interval bounds on the cumulative distribution function (cdf), (2) interval bound on the mean and variance, and (3) a collection of distribution families:

\begin{enumerate}
  \item $\begin{aligned}[t]
    \underline{F}(x) \leq F(x) \leq \overline{F}(x) \\%&&&& \text{(Bounds on the cdf)} \\
  \end{aligned}$
  \item $\begin{aligned}[t]
    \mu &\in [\underline{ \mu }, \overline{ \mu }]  \\%&&&& \text{(bounds on mean and variance)}\\
    \sigma^2 &\in [\underline{\sigma}^2 , \overline{\sigma}^2]
  \end{aligned}$
  \item $\begin{aligned}[t]
      F \in \bold{F} \\%&&&&&&& \text{(distribution family)} \\
  \end{aligned}$
  \end{enumerate}

\noindent That is, a random variable is a member of a p-box if its cdf $F$ falls within the cdf bounds of the p-box $F(x) \in [\underline{F}(x), \overline{F}(x)]$ for all x, its moments are inside the interval moments of the p-box, and it belongs to a family of distribution functions (e.g. normal, uniform) considered by the p-box. Some of the constraints may be missing. For example, if the distribution family is unknown then the set is defined solely from the cdf and moment bounds. Such p-boxes are sometimes called non-parametric, since its members do not have to belong to any particular class of distribution. Some constraints may also be inferred from others. For example, the interval moments may be bounded from the cdf bounds, and cdf may be bounded from moment information (explored further in section \ref{section:Moments}).

All of a p-box's probabilistic quantities are intervals. The cdf of a p-box may be found by:

\begin{equation*}
  [\underline{F}(x), \overline{F}(x)] ,
\end{equation*}

\noindent a sample of the p-box may be drawn using the inverse cdfs:

\begin{equation*}
  [\underline{F}^{-1}(\alpha), \overline{F}^{-1}(\alpha)]
\end{equation*}

\noindent where $\alpha \sim U(0,1)$ is a sample from a uniform distribution, and the probability measure\footnote{The probability measure is a function which returns the probability that the random variable is in some set} on some interval $U = [a, b]$ is bounded as follows:

\begin{align*}
  \underline{\mathbb{P}}(U) &= \text{max}(0, \underline{F}(b) - \overline{F}(a)) \\ 
  \overline{\mathbb{P}}(U)  &= \overline{F}(b) - \underline{F}(a) ,
\end{align*}

where the max operator is required when $\underline{F}(b) < \overline{F}(a)$. Note the same can be achieved by using the standard formula for finding the measure from the cdf, $F(b)- F(a)$, and using interval arithmetic.

\begin{figure*}[htp]

  \centering
  \includegraphics[width=.3\textwidth]{../examples/JuliaCon/fig1/fig1_dist.pdf}\hfill
  \includegraphics[width=.3\textwidth]{../examples/JuliaCon/fig1/fig1_interval.pdf}\hfill
  \includegraphics[width=.3\textwidth]{../examples/JuliaCon/fig1/fig1_pbox.pdf}

  \caption{A precise distribution, an interval and a probability box}
  \label{fig:figure3}

\end{figure*}

P-boxes generalise precise distributions and intervals in the following way. A distribution is a p-box with a precise cdf and moments, i.e. when:

\begin{align*}
  \underline{F}(x) &= \overline{F}(x), \\ 
  \underline{\mu}  &= \overline{\mu}, \\ 
  \underline{\sigma}^2 &= \overline{\sigma}^2
\end{align*}

\noindent and an interval $[a,b]$ is a p-box whose bounds are step functions:

\begin{align*}
    \underline{F}(x) &= \epsilon_{b}(x)\\
    \overline{F}(x) &= \epsilon_{a}(x) ,
\end{align*}

\noindent where $\epsilon_k$ is:

\begin{equation*}
   \epsilon_k(x) = \begin{cases} 0 &\text{when } x < k \\ 1 &\text{when } x \geq k \end{cases}
\end{equation*}

Moreover, theoretical bounds on the mean and variance of an interval can be found [CITE SCOTT]:

\begin{align*}
  \mu &\in [a, b] \\
  \sigma^2 &\in [0, (b - a)^{2/4}]
\end{align*}

That is, it is not possible to find a distribution whose range is in $[a, b]$ and whose variance is greater than $(b-a)^{2/4}$. The lower bound on the variance is zero, since scalars are also included in the interval. Under this definition of an interval, a random sample will always be the interval $[a, b]$, the cdf returns 0 when $x < a$, the interval $[0,1]$ when $a <= x < b$, and $1$ when $b <= x$. Further, the probability measure of the interval $X = [a, b]$ will be:

\begin{equation*}
  \mathbb{P}_{X}(U) = \begin{cases}
    0 & \text{when } U \cap [a,b] = \emptyset \\
    1 & \text{when } [a,b] \subseteq U  \\
    [0, 1] & \text{otherwise }\\
  \end{cases}
\end{equation*}

\noindent That is, if the set $U$ does not intersect the interval $X$, $\mathbb{P}_{X}=0$, i.e. $U$ certainly does not contain any of the random variables. If the X is fully contained in $U$, $\mathbb{P}_{X}=1$, i.e. $U$ certainly contains all of the random variables. And finally if $U$ intersects (but does not contain) $X$, the probability measure is the vacuous probability interval $\mathbb{P}_{X} = [0, 1]$, i.e. we are completely uncertain about the containment. Note that the $P_{X} = \{0, 1, [0,1]\}$ is due to the interval bounds being step function. Generally p-boxes may yield any probability interval.

\subsection{Outer representations of p-boxes}

An important feature of probability bounds analysis is how distribution functions and p-boxes are represented. Generally, analytical solutions for cdfs are not readily available, for example the normal distribution's cdf can only be found by integrating the density, usually done with Gaussian quadrature. However, even if the functions were available analytically, when the variables are used in an arithmetic operation the output distribution will not necessarily belong to the same family. Therefore we require a representation of these continuous functions which is not dependent on the distribution family (for example Polynomial chaos expansion), and ideally is robust.

@@

\subsubsection{Partial moment information} %\hfill \break
\label{section:Moments}
%It is well known that a random variables distribution function cannot be

@@

\section{P-box arithmetic}

@@

The long term goal of such a framework is to create a fully uncertain programming language, where any computer variable may be represented as an interval, distribution, p-box or other uncertain quantity. Such a framework would allow for uncertain extensions of deterministic functions to be computed in an automatic, rigorous and tight fashion. In this section we argue why Julia is an ideal target language for such a framework, and discuss some of the remaining theoretical tasks required to make such a goal a reality.

\section*{Acknowledgements}

The authors would like to thank the support of this work from the Engineering Physical Sciences Research Council (EPSRC) iCase studentship award. This research is funded by the EPSRC with grant number EP/R006768/1, which is acknowledged for its funding and support. This work has been carried out within the frame- work of the EUROfusion Consortium and has received funding from the Euratom research and training programme 2014–2018 and 2019–2020 under grant agreement number 633053. The views and opinions expressed herein do not necessarily reflect those of the European Commission
